\documentclass[11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{longtable}

\geometry{letterpaper, margin=1in}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{red},
  breaklines=true,
  columns=fullflexible,
  keepspaces=true,
  language=C++
}

\title{\textbf{VSEPR-SIM Automated Report} \\
\Large Report Automation Pipeline Output}
\author{VSEPR-SIM Reporting System v3.0.0}
\date{\today}

\begin{document}
\maketitle
\tableofcontents
\newpage

% ============================================================================
\section{Introduction}

This document is produced by the VSEPR-SIM automated reporting pipeline.
All values are deterministic and traceable to source computations.
No hidden state is involved at any layer.

% ============================================================================
\section{Report Modules}

The C++ report-generation functions available to users are listed below.
Each function produces explicit, inspectable output suitable for research
documentation.

\begin{table}[h]
\centering
\begin{tabular}{lll}
\toprule
\textbf{Module} & \textbf{Function} & \textbf{Output} \\
\midrule
FIRE Report         & \texttt{bead\_fire\_report\_md()}       & Markdown \\
FIRE File Export    & \texttt{write\_bead\_fire\_report()}     & \texttt{.md} \\
FIRE Summary        & \texttt{bead\_fire\_summary()}           & Console \\
CSV Trajectory      & \texttt{write\_bead\_fire\_csv()}        & \texttt{.csv} \\
Mapping Report      & \texttt{write\_mapping\_report()}        & \texttt{.md} \\
Excel XML Export    & \texttt{export\_excel\_xml()}            & \texttt{.xml} \\
SolidWorks Export   & \texttt{export\_solidworks\_sldcrv()}    & \texttt{.sldcrv} \\
Snapshot \& Graph   & \texttt{SnapshotGraphCollector}          & CSV / Markdown \\
Atomistic FIRE      & \texttt{fire\_report\_md()}              & Markdown \\
\bottomrule
\end{tabular}
\caption{User-side report functions.}
\end{table}

% ============================================================================
\section{Automation Pipeline}

The pipeline proceeds as follows:

\begin{enumerate}
  \item C++ simulation runs and calls report functions to write artifacts.
  \item \texttt{reporting/generate\_report.py} scans the output directory.
  \item Trajectory CSVs are parsed and matplotlib figures are generated.
  \item A consolidated Markdown report is assembled.
  \item A manifest JSON is written for downstream tooling.
\end{enumerate}

% ============================================================================
\section{Mathematical Framework}

FIRE (Fast Inertial Relaxation Engine) velocity-Verlet integration:

\begin{equation}
  \mathbf{F}_i = -\nabla_{\mathbf{r}_i} U_{\text{total}}(\{\mathbf{r}\}, \{\mathbf{q}\})
\end{equation}

\begin{equation}
  \mathbf{r}_{i,t+1} = \mathbf{r}_{i,t} + \Delta t \, \mathbf{v}_{i,t}
\end{equation}

\begin{equation}
  \mathbf{v}_{i,t+1} = (1-\alpha)\mathbf{v}_{i,t}
    + \alpha\,\frac{\mathbf{F}_{i,t}}{\|\mathbf{F}_t\|}\,\|\mathbf{v}_{i,t}\|
\end{equation}

Convergence criteria:
\begin{equation}
  \|\mathbf{F}\|_{\text{RMS}} < \varepsilon_F
  \quad \lor \quad
  \frac{|U_t - U_{t-1}|}{N_{\text{bead}}} < \varepsilon_U
\end{equation}

% ============================================================================
\section{Running Reports}

\subsection{Single Report}
\begin{lstlisting}[language=bash]
./scripts/run_report.sh
\end{lstlisting}

\subsection{Batch Reports}
\begin{lstlisting}[language=bash]
./scripts/batch_reports.sh [num_runs]
\end{lstlisting}

\subsection{Demo Mode}
\begin{lstlisting}[language=bash]
python reporting/generate_report.py --demo --out outputs/demo_report
\end{lstlisting}

% ============================================================================
\section{Output Properties}

All outputs satisfy the anti-black-box design philosophy:

\begin{itemize}
  \item \textbf{Explicit} --- every value maps to a specific computation.
  \item \textbf{Inspectable} --- Markdown and CSV are human-readable.
  \item \textbf{Deterministic} --- same input yields identical output.
  \item \textbf{Modular} --- each report module operates independently.
  \item \textbf{Traceable} --- every metric links to a source step index.
\end{itemize}

\vfill
\noindent\rule{\textwidth}{0.4pt}
\small\textit{Generated by VSEPR-SIM automated reporting pipeline.}

\end{document}
