\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{listings}
\usepackage{xcolor}

\geometry{letterpaper, margin=1in}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{red},
  breaklines=true,
  columns=fullflexible,
  keepspaces=true,
  language=C++
}

\title{\textbf{Formation Engine Methodology} \\
\Large Phase 1 --- Core Kernel Revalidation Report}
\author{Formation Engine Development Team}
\date{Version 0.1 --- Live Document (Updates With Each Phase)}

\begin{document}
\maketitle

\section{Purpose}

This document records the results of deterministic revalidation of the core
atomistic kernel, following the phased development plan.  It is a live
document: new sections are appended as each phase completes and its
verification program confirms pass/fail status.

Each phase produces a self-testing C++ executable that prints pass/fail for
every check.  No external test frameworks, scripts, or JSON configurations
are involved.

\section{Phase 1 --- Core Kernel Audit}

\textbf{Executable:} \texttt{apps/phase1\_kernel\_audit.cpp}

\textbf{Date:} As built from current codebase.

\subsection{Scope}

Phase~1 revalidates the lowest layer of the system: single-point energy
evaluation and force computation on the simplest possible system (an Ar
dimer).  The three audit objectives are:

\begin{enumerate}
\item \textbf{Step~1.1:} Verify that the energy ledger is self-consistent.
\item \textbf{Step~1.2:} Verify that evaluation is bitwise deterministic.
\item \textbf{Step~1.3:} Verify that forces are finite, physically directional,
      and consistent with the energy surface.
\end{enumerate}

\subsection{Active Energy Model}

The current kernel evaluates the Lennard-Jones 12--6 potential with UFF
parameters and Lorentz--Berthelot combining rules:
\begin{equation}
U_{\mathrm{LJ}}(r) = 4\varepsilon_{ij}
    \left[\left(\frac{\sigma_{ij}}{r}\right)^{12}
    - \left(\frac{\sigma_{ij}}{r}\right)^{6}\right]
    \cdot S(r;\, r_{\mathrm{on}},\, r_c)
\end{equation}
where $S$ is a quintic switching function active between $0.9\,r_c$ and
$r_c$.  Implementation: \texttt{atomistic/models/lj\_coulomb.cpp}.

\subsubsection{Declared Limitation: Coulomb Disabled}

The Coulomb term is implemented but currently zeroed out in the kernel
(\texttt{U\_coul = 0.0}, \texttt{F\_coul\_r = 0.0}).  This was disabled
after a systematic instability was observed during NaCl MD integration.
The root cause was identified as an integrator--Coulomb coupling issue
that has not yet been resolved.

\textbf{Consequence:} All ionic and polar systems are currently evaluated
using van~der~Waals interactions only.  Phase~1 records this as a known
limitation.  Re-enabling Coulomb is deferred until a dedicated
Coulomb-stability audit is performed (planned for Phase~2 extension).

The Coulomb interface and charge arrays (\texttt{State::Q}) remain intact
so that re-enabling requires only uncommenting two lines in
\texttt{lj\_coulomb.cpp} and passing the stability audit.

\subsection{Results}

All 11 checks passed.

\begin{center}
\begin{tabular}{llll}
\toprule
Step & Check & Metric & Status \\
\midrule
1.1 & \texttt{total() == sum of components}
    & $|\Delta| < 10^{-15}$ & PASS \\
1.1 & All energy terms finite
    & \texttt{isfinite()} & PASS \\
1.1 & Coulomb is zero (declared limitation)
    & $U_{\mathrm{Coul}} = 0$ & PASS \\
\midrule
1.2 & Identical inputs $\to$ identical energy
    & bitwise & PASS \\
1.2 & Identical inputs $\to$ identical forces
    & bitwise & PASS \\
\midrule
1.3a & Forces and energies finite $\forall\, r \in [2.5, 10.0]$\,\AA
    & sweep, $\Delta r = 0.1$ & PASS \\
1.3b & $U(r_{\min}) < 0$ (bound state)
    & $U = -0.238$\,kcal/mol & PASS \\
1.3b & $|F(r_{\min})| < 0.01$
    & $4.4 \times 10^{-4}$ & PASS \\
1.3b & Repulsive at $r < r_{\min}$
    & $F_x < 0$ & PASS \\
1.3b & Attractive at $r > r_{\min}$
    & $F_x > 0$ & PASS \\
1.3c & $F_x = -\partial U/\partial x_0$ (finite diff.)
    & $|\mathrm{err}| = 9.5 \times 10^{-12}$ & PASS \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Measured Constants}

From the Ar dimer scan:
\begin{center}
\begin{tabular}{lrl}
\toprule
Quantity & Value & Source \\
\midrule
$r_{\min}$ & 3.816\,\AA & grid search, $\Delta r = 0.001$\,\AA \\
$U(r_{\min})$ & $-0.238$\,kcal/mol & single-point at $r_{\min}$ \\
$|F(r_{\min})|$ & $4.4 \times 10^{-4}$\,kcal/mol/\AA & analytic \\
Force--energy error & $9.5 \times 10^{-12}$\,kcal/mol/\AA & central diff.\ $\delta = 10^{-5}$ \\
\bottomrule
\end{tabular}
\end{center}

These are consistent with UFF Ar parameters ($\varepsilon = 0.238$\,kcal/mol,
$\sigma = 3.40$\,\AA, $r_{\min} = 2^{1/6}\sigma = 3.816$\,\AA).

\subsection{Conclusion}

The LJ nonbonded kernel is self-consistent, deterministic, and
force--energy coherent.  The Coulomb term is disabled and declared as a
known limitation.  The energy ledger bookkeeping is correct.

\textbf{Phase~1 status: PASS.}

%% ====================================================================
\section{Phase 2 --- Structural Energy Physics}
%% ====================================================================

\textbf{Executable:} \texttt{apps/phase2\_structural\_energy.cpp}

\subsection{Scope}

Phase~2 evaluates the kernel on chemically meaningful reference systems by
routing them through the \emph{full end-user pipeline}: element symbols and
Cartesian coordinates are packed into \texttt{vsepr::io::XYZMolecule},
parsed by \texttt{atomistic::parsers::from\_xyz()}, then evaluated by the
LJ model.  This is the same path a user's \texttt{.xyz} file would take.

\subsection{Bug Fixed During Phase 2}

\textbf{Parser type-map bug.}
The XYZ parser (\texttt{atomistic/parsers/xyz\_parser.cpp}) was assigning
\texttt{State::type[i]} as a sequential index (0, 1, 2, \ldots) rather
than the atomic number~$Z$.  Consequence: for H$_2$O, hydrogen was mapped
to type~0 (undefined) and oxygen to type~1 (hydrogen's LJ parameters).
Every multi-element molecule was using wrong LJ parameters.

\textbf{Fix:} Replaced the sequential type map with a direct call to
\texttt{get\_atomic\_number()} so that \texttt{type[i] = Z_i}.  Verified:
O~$\to$~8, H~$\to$~1, C~$\to$~6, Ar~$\to$~18.

\subsection{Declared Finding: Bonded-Pair Exclusions Not Yet Applied}

The LJ model evaluates \emph{all} atom pairs including covalently bonded
ones.  At bond-length separations ($r \approx 0.7$--$1.1$\,\AA), atoms are
deep inside the LJ repulsive wall ($r \ll \sigma$), producing energies of
$10^5$--$10^6$\,kcal/mol for small molecules.  This is mathematically
correct but physically nonsensical for bonded pairs.

\textbf{Consequence:} Molecular energies from the nonbonded model alone
are dominated by bonded-pair LJ repulsion.  Meaningful molecular energetics
require either bonded-pair exclusion lists in the nonbonded model, or a
composite model that subtracts bonded-pair LJ contributions.

This is recorded as a declared next step for Phase~3 (relaxation with
bonded terms).

\subsection{Results}

All 15 checks passed.

\begin{center}
\begin{tabular}{lll}
\toprule
Step & Check & Status \\
\midrule
2.1 & H$_2$, H$_2$O, CH$_4$, Ar$_2$: all finite & PASS \\
2.1 & H$_2$O parser: O $\to$ type=8, H $\to$ type=1 & PASS \\
2.1 & CH$_4$ parser: C $\to$ type=6, H $\to$ type=1 & PASS \\
\midrule
2.2 & H$_2$O: energy bitwise identical across runs & PASS \\
2.2 & H$_2$O: forces bitwise identical across runs & PASS \\
\midrule
2.3 & H$_2$: compressed energy $>$ equilibrium energy & PASS \\
2.3 & H$_2$O: stretching one O--H changes total energy & PASS \\
\midrule
2.4 & Ar$_2$ sweep finite $\forall\, r \in [3,8]$\,\AA & PASS \\
2.4 & Ar$_2$ max $|\Delta U|$ per 0.01\,\AA\ step $< 1$\,kcal/mol & PASS \\
\midrule
2.5 & H$_2$O: 9 force components match $-\partial U/\partial x_i$ (rel tol $10^{-6}$) & PASS \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Reference Energies (nonbonded LJ only, no exclusions)}

\begin{center}
\begin{tabular}{lrrl}
\toprule
System & $N$ & $U_{\mathrm{vdW}}$\,(kcal/mol) & Note \\
\midrule
H$_2$   & 2 & $2.18 \times 10^{6}$ & deep in LJ wall ($r = 0.74$\,\AA, $\sigma_\mathrm{H} \approx 2.9$\,\AA) \\
H$_2$O  & 3 & $7.68 \times 10^{5}$ & bonded pairs dominate \\
CH$_4$  & 5 & $8.25 \times 10^{5}$ & bonded pairs dominate \\
Ar$_2$  & 2 & $-0.224$             & physical: near LJ minimum \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Conclusion}

The end-user pipeline (XYZ parse $\to$ State $\to$ eval) is deterministic,
force--energy coherent, and produces correct atomic-number type assignments
after the parser fix.  Ar$_2$ energies are physically meaningful.  Molecular
energies are dominated by bonded-pair LJ repulsion, confirming the need for
bonded-pair exclusions (Phase~3).

\textbf{Phase~2 status: PASS.}

%% ====================================================================
\section{Phase 3 --- Relaxation and Minimization}
%% ====================================================================

\textbf{Executable:} \texttt{apps/phase3\_relaxation.cpp}

\subsection{Scope}

Phase~3 tests the FIRE energy minimizer on distorted structures routed
through the full end-user pipeline including bond detection.  This is the
first phase to exercise the 1--2 bonded-pair exclusion code added to the
nonbonded model during this phase.

\subsection{Code Changes}

\subsubsection{1--2 Bonded-Pair Exclusion in LJ Model}

The \texttt{LJCoulomb::eval()} method now builds an exclusion set from
\texttt{State::B} at the start of each evaluation.  Atom pairs connected by
a bond edge are skipped during the nonbonded loop.  This eliminates the
unphysical $10^5$--$10^6$\,kcal/mol repulsion that Phase~2 identified for
bonded atoms at covalent distances.

\textbf{Implementation:} A \texttt{std::unordered\_set<uint64\_t>} hashes
$(i,j)$ pairs from \texttt{State::B} using the key
$(\min(i,j) \ll 32) \,|\, \max(i,j)$.  Each pair in the inner loop is
checked against this set before evaluation.

\subsubsection{Verification}

H$_2$ at $r = 0.74$\,\AA:
\begin{center}
\begin{tabular}{lrl}
\toprule
Configuration & $U_{\mathrm{vdW}}$\,(kcal/mol) & Behaviour \\
\midrule
Without bond & $2.18 \times 10^{6}$ & deep LJ repulsion \\
With bond    & $0.000$ & pair excluded \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Declared Finding: Bonded Model Not Yet Composed}

With 1--2 exclusions active, bonded atoms have \emph{no interaction}:
neither LJ (excluded) nor harmonic spring (\texttt{BondedModel} not yet
composed with \texttt{LJCoulomb}).  Molecules relax under nonbonded
interactions between 1--3+ pairs only.

\textbf{Consequence:} The Ar$_3$ nonbonded cluster converges to tight
tolerance.  H$_2$O and CH$_4$ show energy decrease and bond retention but
cannot reach tight force convergence without bonded terms.  Full molecular
relaxation requires composing \texttt{BondedModel} with \texttt{LJCoulomb}
(deferred to Phase~4).

\subsection{Results}

All 17 checks passed.

\begin{center}
\begin{tabular}{lll}
\toprule
Step & Check & Status \\
\midrule
3.6 & H$_2$ no bonds: $U > 10^5$ (LJ repulsion) & PASS \\
3.6 & H$_2$ with bonds: $U = 0$ (pair excluded) & PASS \\
\midrule
3.1 & Ar$_3$: energy decreased ($-0.43 \to -0.71$) & PASS \\
3.2 & Ar$_3$: $F_{\mathrm{rms}} = 7.9 \times 10^{-5}$ (converged) & PASS \\
3.4 & Ar$_3$: equilateral, $d = 3.816$\,\AA\ (LJ minimum) & PASS \\
\midrule
3.1 & H$_2$O: energy decreased ($72.6 \to 0.83$) & PASS \\
3.3 & H$_2$O: bond count preserved (2) & PASS \\
3.1 & CH$_4$: energy decreased ($99.2 \to 8.1$) & PASS \\
3.3 & CH$_4$: bond count preserved (4) & PASS \\
\midrule
3.5 & Ar$_3$: relaxed energy bitwise deterministic & PASS \\
3.5 & Ar$_3$: relaxed positions bitwise deterministic & PASS \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Measured Constants}

Ar$_3$ cluster after FIRE relaxation:
\begin{center}
\begin{tabular}{lrl}
\toprule
Quantity & Value & Note \\
\midrule
All pair distances & 3.816\,\AA & exact LJ $r_{\min}$ \\
$U_{\mathrm{total}}$ & $-0.714$\,kcal/mol & $= 3 \times (-0.238)$ (3 pairs) \\
$F_{\mathrm{rms}}$ & $7.9 \times 10^{-5}$ & after 6878 FIRE steps \\
\bottomrule
\end{tabular}
\end{center}

The Ar$_3$ total energy is exactly three times the Ar$_2$ dimer energy,
confirming additive pairwise LJ and correct exclusion logic (no bonds in
the Ar system, so nothing is excluded).

\textbf{Phase~3 status: PASS.}

%% ====================================================================
\section{Phase 4 --- Formation Priors}
%% ====================================================================

\textbf{Executable:} \texttt{apps/phase4\_formation\_priors.cpp}

\subsection{Scope}

Phase~4 verifies that the crystal preset library and formation-prior
infrastructure are correctly separated from the physical kernel.  The
test exercises the crystal end-user pipeline:
\[
\texttt{presets::X()} \;\to\; \texttt{UnitCell}
    \;\to\; \texttt{to\_state()} \;\to\; \texttt{model->eval()} \;\to\; E
\]

Eight crystal presets are verified across three checks: geometric
correctness (atom count, lattice constant, space group number),
\texttt{State} well-formedness (PBC enabled, all arrays sized),
and single-point energy finiteness and determinism.

\subsection{Crystal Preset Library}

The atomistic kernel ships a comprehensive crystal preset library
(\texttt{atomistic/crystal/unit\_cell.hpp}) with the following categories:

\begin{center}
\begin{tabular}{llrl}
\toprule
Preset & Structure & $N_{\mathrm{basis}}$ & $a$\,(\AA) \\
\midrule
FCC Al  & Fm$\overline{3}$m (225) & 4 & 4.05 \\
BCC Fe  & Im$\overline{3}$m (229) & 2 & 2.87 \\
NaCl    & Fm$\overline{3}$m (225) & 8 & 5.64 \\
Diamond Si & Fd$\overline{3}$m (227) & 8 & 5.43 \\
FCC Cu  & Fm$\overline{3}$m (225) & 4 & 3.61 \\
FCC Au  & Fm$\overline{3}$m (225) & 4 & 4.08 \\
CsCl    & Pm$\overline{3}$m (221) & 2 & 4.12 \\
MgO     & Fm$\overline{3}$m (225) & 8 & 4.21 \\
\bottomrule
\end{tabular}
\end{center}

Plus $\sim$25 additional presets (fluorites, spinels, perovskites,
garnets, pyrochlores, actinide oxides) for future Phase~5 validation.

\subsection{Supercell Construction}

The 2$\times$2$\times$2 Al FCC supercell produces 32~atoms with
$L_x = 8.10$\,\AA\ $= 2a$ and PBC enabled.  Per-atom energy differs
significantly from the unit cell because the cell dimension
($8.1$\,\AA) is smaller than $2 r_c$ ($20$\,\AA), so minimum-image
convention undersamples the periodic LJ sum.  This is a known
finite-size effect, not a construction bug.

\subsection{Results}

All 52 checks passed (8 presets $\times$ 3 geometry checks $+$ 8 State
checks $+$ 16 energy checks $+$ 4 supercell checks).

\textbf{Phase~4 status: PASS.}

%% ====================================================================
\section{Phase 5 --- Crystal Structure Validation}
%% ====================================================================

\textbf{Executable:} \texttt{apps/phase5\_crystal\_validation.cpp}

\subsection{Scope}

Phase~5 emits crystal supercells from the preset library, measures
nearest-neighbor distances and coordination numbers under PBC, relaxes
with FIRE, and verifies structural preservation.

\subsection{Geometric Verification}

\begin{center}
\begin{tabular}{lrrlrrl}
\toprule
Crystal & $N$ & $d_{\mathrm{nn}}$\,(\AA) & Expected & CN & Expected & Status \\
\midrule
FCC Al  & 108 & 2.864 & $a/\!\sqrt{2}=2.864$ & 12 & 12 & PASS \\
BCC Fe  & 128 & 2.485 & $a\sqrt{3}/2=2.485$  &  8 &  8 & PASS \\
NaCl    &  64 & 2.820 & $a/2=2.820$          &  6 &  6 & PASS \\
Diamond Si & 64 & 2.351 & $a\sqrt{3}/4=2.351$ & 4 & 4 & PASS \\
\bottomrule
\end{tabular}
\end{center}

All nearest-neighbor distances and coordination numbers match
crystallographic expectations exactly.

\subsection{Relaxation Preservation}

BCC Fe 4$\times$4$\times$4 (128 atoms, $L = 11.48$\,\AA):
the structure is already at a local minimum under the LJ model.  FIRE
converges in 2~steps with $F_{\mathrm{rms}} = 5.8 \times 10^{-3}$.
Nearest-neighbor distance is preserved to better than $10^{-4}$\,\AA.

\subsection{Energy Convergence}

FCC Cu per-atom energy: 3$\times$3$\times$3 ($N=108$) vs
4$\times$4$\times$4 ($N=256$) differ by $0.016$\,kcal/mol/atom (0.36\%),
confirming convergence once the cell exceeds $2 r_c$.

\textbf{Phase~5 status: PASS.}

%% ====================================================================
\section{Phases 6/7 --- Deterministic Verification Session Ladder}
%% ====================================================================

\textbf{Executable:} \texttt{apps/phase6\_7\_verification\_sessions.cpp}

\subsection{Scope}

Phases~6 and~7 build and run the full verification session ladder.  Each
named session:
\begin{itemize}
\item runs a reproducible computation with explicit per-check thresholds
\item writes \texttt{report.txt} (metrics, energy tables, geometry) and
      \texttt{status.txt} (\texttt{PASS}/\texttt{FAIL}) to
      \texttt{verification/<group>/<id>/}
\item passes mock end-user inputs through the full pipeline:
      \texttt{XYZMolecule} $\to$ \texttt{detect\_bonds} $\to$
      \texttt{parsers::from\_xyz} $\to$ \texttt{model->eval} $\to$ FIRE
\end{itemize}

\subsection{Session Ladder}

\subsubsection{Block A --- Structural Energy}

\begin{center}
\begin{tabular}{lll}
\toprule
Session & Description & Status \\
\midrule
\texttt{se\_001} & H$_2$ LJ distance sweep, r = 3--8\,\AA\ (nonbonded) & PASS \\
\texttt{se\_002} & H$_2$O O--H stretch, r = 0.8--3.0\,\AA\ + determinism & PASS \\
\texttt{se\_003} & CH$_4$ C--H stretch, d = 0.8--2.0\,\AA\ + compression test & PASS \\
\texttt{se\_004} & Ar$_2$ pair sweep, r = 3--10\,\AA\ + force direction & PASS \\
\bottomrule
\end{tabular}
\end{center}

\subsubsection{Block B --- Relaxation}

\begin{center}
\begin{tabular}{lll}
\toprule
Session & Description & Status \\
\midrule
\texttt{rx\_001} & Ar$_3$ equilateral cluster FIRE from 4.5\,\AA\ & PASS \\
\texttt{rx\_002} & H$_2$O distorted (130°) FIRE, bond retention & PASS \\
\texttt{rx\_003} & BCC Fe $3\!\times\!3\!\times\!3$ FIRE under PBC & PASS \\
\bottomrule
\end{tabular}
\end{center}

\subsubsection{Block C --- Crystal}

\begin{center}
\begin{tabular}{llll}
\toprule
Session & Preset & Geometry checks & Status \\
\midrule
\texttt{xtal\_001} & FCC Al  & nn, CN, relax, energy & PASS \\
\texttt{xtal\_002} & BCC Fe  & nn, CN, relax, energy & PASS \\
\texttt{xtal\_003} & NaCl    & nn, CN, relax, energy & PASS \\
\texttt{xtal\_004} & Diamond Si & nn, CN, energy      & PASS \\
                   &            & \textit{relax skipped: nn $\ll$ $\sigma_{\mathrm{LJ}}$} & \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Findings}

\textbf{Bond detection discontinuity.}  The se\_001 H$_2$ sweep originally
ran from 0.5\,\AA\ with bond detection enabled.  At the covalent detection
threshold (${\sim}0.74$\,\AA) the H--H pair transitions from ``bonded
(excluded, $U=0$)'' to ``nonbonded (deep LJ repulsion)'', creating a step
discontinuity of ${\sim}10^6$\,kcal/mol.  This is physically correct
behaviour but not useful as a smoothness benchmark.  The sweep was
re-scoped to r\,=\,3--8\,\AA\ (nonbonded, no bond detection), which is the
physically relevant LJ region.

\textbf{Covalent crystals.}  Diamond Si (nn\,=\,2.35\,\AA,
$\sigma_{\mathrm{Si}}$\,=\,3.83\,\AA) and any crystal with
nn\,$<$\,2.5\,\AA\ are flagged at runtime and their FIRE relaxation steps
are skipped with a logged note.  This avoids FIRE divergence under forces
of ${\sim}10^3$\,kcal/mol/\AA\ and documents the requirement for bonded
terms in covalent systems.

\subsection{Artifact Structure}

\begin{verbatim}
verification/
  status.txt                  (master: Sessions/Passed/Failed/Result)
  structural_energy/
    se_001_h2_distance_sweep/ (report.txt, status.txt)
    se_002_h2o_oh_stretch/
    se_003_ch4_bond_stretch/
    se_004_ar2_pair_sweep/
  relaxation/
    rx_001_ar3_fire/
    rx_002_h2o_fire/
    rx_003_bcc_fe_crystal/
  crystals/
    xtal_001_fcc_al/
    xtal_002_bcc_fe/
    xtal_003_nacl/
    xtal_004_diamond_si/
\end{verbatim}

\textbf{Phases~6/7 status: PASS.}  43 checks, 0 failures.  All 11 session
\texttt{status.txt} files read \texttt{PASS}.

%% ====================================================================
\section{Phases 8/9/10 --- Deferred Physics, Architecture, Environment}
%% ====================================================================

\textbf{Executable:} \texttt{apps/phase8\_9\_10\_environment.cpp}

\subsection{Phase 8: Cleanly Defer Missing Physics}

The \texttt{atomistic/core/physics\_status.hpp} header is the
authoritative record of which physics terms are active, incomplete, or
deferred.  The full registry:

\begin{center}
\begin{tabular}{ll}
\toprule
Term & State \\
\midrule
LJ 12-6 nonbonded          & ACTIVE \\
Harmonic bonds              & PARTIAL (not composed) \\
Harmonic angles             & PARTIAL (not composed) \\
Periodic torsions           & PARTIAL (not composed) \\
Coulomb electrostatics      & DEFERRED \\
Dipole--dipole interactions & DEFERRED \\
Induced dipoles / SCF       & PARTIAL \\
Environment screening       & ABSENT \\
PBC / MIC                   & ACTIVE \\
FIRE minimisation           & ACTIVE \\
Velocity Verlet NVE         & ACTIVE \\
Langevin NVT                & ACTIVE \\
\bottomrule
\end{tabular}
\end{center}

Coulomb and dipole interactions are deferred with interfaces preserved,
not deleted.  The polarization SCF solver exists and passes internal
consistency checks but is excluded from the default eval chain until
Coulomb is stabilised.

\subsection{Phase 9: Layered Architecture}

The \texttt{atomistic/core/layer\_manifest.hpp} header documents five
engine layers:

\begin{enumerate}
\item \textbf{Formation Priors} --- VSEPR seeds, crystal presets, bond detection (partial)
\item \textbf{Deterministic Kernel} --- LJ+PBC, bonded model, integrators (partial)
\item \textbf{Verification} --- Phases 1--7, session artifacts (complete)
\item \textbf{Analysis \& Ranking} --- Crystal metrics (partial)
\item \textbf{Bridge \& Artifact} --- EngineAdapter, SceneDocument (partial)
\end{enumerate}

\subsection{Phase 10: Environmental Context}

\texttt{atomistic/core/environment.hpp} introduces the
\texttt{EnvironmentContext} struct with three medium classes:

\begin{itemize}
\item \textbf{NearVacuum} --- $\varepsilon_r = 1$, no screening
\item \textbf{DryCondensed} --- $\varepsilon_r \approx 2$--$5$, bulk solid/film
\item \textbf{Solution} --- $\varepsilon_r \gg 1$ (water: 78.4),
      ionic strength, Debye screening
\end{itemize}

\texttt{EnvironmentContext} is wired into \texttt{ModelParams::env}
(kernel side) and \texttt{KernelRequest::env} (bridge side).

Verified: \texttt{coulomb\_scale()} = $1/\varepsilon_r$,
Debye length $\lambda_D = 9.61$\,\AA\ at $I = 0.1$\,mol/L in water.
LJ energy on a neutral Ar dimer is unaffected by environment (correct:
Coulomb is deferred, charges are zero).

\textbf{Bridge wiring:} \texttt{EmitCrystal} KernelOp now implemented
in \texttt{EngineAdapter.cpp}.  The desktop can emit any of the 8+
crystal presets with configurable supercell dimensions.  The environment
context from \texttt{KernelRequest} is automatically propagated to
\texttt{ModelParams::env} on every kernel call.

\subsection{Results}

All 29 checks passed.

\textbf{Phases~8/9/10 status: PASS.}

%% ====================================================================
\section{Phases 11/12/13 --- Composite Model, Bridge, Final Regression}
%% ====================================================================

\textbf{Executable:} \texttt{apps/phase11\_12\_13\_final.cpp}

\subsection{Phase 11: Compose BondedModel + LJCoulomb}

A new \texttt{CompositeModel} class (\texttt{atomistic/models/composite.hpp})
evaluates multiple \texttt{IModel} instances and sums forces and energy.
The factory \texttt{create\_composite\_model(s)} composes:
\begin{enumerate}
\item \texttt{BondedModel} (harmonic bonds, angles, torsions from \texttt{State::B})
\item \texttt{LJCoulomb} (with 1--2 exclusions)
\end{enumerate}

Verified: H$_2$O and CH$_4$ composite evaluations produce nonzero
$U_{\mathrm{bond}}$ and $U_{\mathrm{angle}}$ alongside $U_{\mathrm{vdW}}$.
Energy decreases under FIRE. Ar dimer through composite infrastructure passes
force--energy consistency at $10^{-6}$.

\textbf{Declared limitation:} The generic bonded parameters set
$r_0 = $ current bond length and $\theta_0 = $ current angle.  These do not
account for nonbonded equilibria, so bonded and LJ forces compete rather
than cooperate.  Tight force convergence on molecules requires matched
force-field parameters (AMBER, CHARMM, etc.) --- a future milestone.
The angle gradient in \texttt{BondedModel} also shows $\sim$7\% discrepancy
versus numeric derivatives, declared for future investigation.

\subsection{Phase 12: Bridge Integration}

\begin{itemize}
\item \texttt{EmitCrystal} KernelOp: implemented in \texttt{EngineAdapter.cpp},
      maps 8 preset names to crystal presets with configurable supercell.
\item \texttt{EnvironmentContext} propagates from \texttt{KernelRequest::env}
      to \texttt{ModelParams::env} on every kernel call.
\item \texttt{SinglePoint} returns full energy decomposition
      ($U_{\mathrm{bond}}, U_{\mathrm{angle}}, U_{\mathrm{vdW}}, \ldots$).
\item Load $\to$ Relax round-trip: energy decreases, bonds preserved.
\end{itemize}

\subsection{Phase 13: Full Regression Sweep}

All prior phase executables re-executed from within Phase~13:

\begin{center}
\begin{tabular}{lrl}
\toprule
Phase & Checks & Exit code \\
\midrule
1 (kernel)     & 11 & 0 \\
2 (structural) & 15 & 0 \\
3 (relaxation) & 17 & 0 \\
4 (presets)    & 52 & 0 \\
5 (crystals)   & 13 & 0 \\
6/7 (sessions) & 43 & 0 \\
8--10 (env)    & 29 & 0 \\
11--13 (this)  & 26 & 0 \\
\midrule
\textbf{Total} & \textbf{206} & \textbf{all 0} \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Phases~11/12/13 status: PASS.  Zero regressions across all phases.}

%% ====================================================================
\section{Phase 14 --- Five-Milestone Implementation}
%% ====================================================================

\textbf{Executable:} \texttt{apps/phase14\_milestones.cpp}

\subsection{M1: Coulomb Re-enablement}

Coulomb interactions re-enabled in \texttt{lj\_coulomb.cpp} with dielectric
screening from \texttt{EnvironmentContext}:
\[
    U_{\mathrm{Coul}}(r) = \frac{k_e}{\varepsilon_r}\,\frac{q_i\,q_j}{r}
    \qquad
    F(r) = \frac{k_e}{\varepsilon_r}\,\frac{q_i\,q_j}{r^2}\,\hat{r}_{ij}
\]
where $k_e = 332.0636\;\mathrm{kcal\!\cdot\!\AA/(mol\!\cdot\!e^2)}$ (AMBER).

Verified: NaCl pair at $r = 2.82\;\mathrm{\AA}$ yields
$U_{\mathrm{Coul}} = -117.75\;\mathrm{kcal/mol}$ (exact: $-k_e/r = -117.75$).
Force--energy consistency $2.8 \times 10^{-10}$ error.
Neutral Ar dimer: $U_{\mathrm{Coul}} = 0$ exactly.

\subsection{M2: UFF-Matched Bonded Parameters}

\texttt{create\_generic\_bonded\_model} now sets:
\begin{itemize}
\item $r_0 = $ current bond length (geometry-matched)
\item $\theta_0 = $ current angle (geometry-matched)
\item $k_b = 664.12\,Z_{\mathrm{eff}}/r_0^2$ (UFF-inspired, element-aware)
\end{itemize}

At the input geometry, $U_{\mathrm{bond}} = U_{\mathrm{angle}} = 0$ exactly.
Distorted H$_2$O relaxes back toward $r_0$ with energy decrease and bond
contraction.

\subsection{M3: Angle Gradient Sign Fix}

The harmonic angle force had a sign error:
\begin{align*}
    \text{Bug:} \quad k &= -2k_\theta(\theta-\theta_0)/\sin\theta \quad\text{(wrong)} \\
    \text{Fix:} \quad k &= +2k_\theta(\theta-\theta_0)/\sin\theta \quad\text{(correct)}
\end{align*}
Derivation: $F_i = -\partial U/\partial\mathbf{r}_i$, where
$\partial\theta/\partial\mathbf{r}_i = (-1/\sin\theta)\,\partial(\cos\theta)/\partial\mathbf{r}_i$,
yields $F_i = +2k_\theta(\theta-\theta_0)/\sin\theta \cdot [\hat{r}_{kj}/(|\mathbf{r}_{ij}||\mathbf{r}_{kj}|) - \cos\theta\,\hat{r}_{ij}/|\mathbf{r}_{ij}|^2]$.

After fix: H$_2$O composite force--energy consistency $5.7\times 10^{-9}$
relative error (was $\sim$7\% before).

\subsection{M4: Crystal/Cluster Relaxation}

Ar$_5$ cluster relaxes through composite model (no bonds $\to$ pure LJ
passthrough). BCC Fe 3$\times$3$\times$3 energy finite at LJ equilibrium.

\textbf{Declared limitation:} covalent crystals (Si diamond) require
1--3 exclusions or reactive potentials (Tersoff/REBO). Second-nearest
Si--Si at 3.84\,\AA\ $<$ $\sigma_{\mathrm{Si}} = 4.30$\,\AA\ causes
unphysical LJ repulsion even with 1--2 exclusions.

\subsection{M5: Dielectric Screening}

\begin{center}
\begin{tabular}{lll}
\toprule
Test & Expected & Result \\
\midrule
$U_{\mathrm{Coul}}^{\mathrm{vac}} / U_{\mathrm{Coul}}^{\mathrm{water}}$ & 78.40 & 78.40 \\
$U_{\mathrm{Coul}}^{\mathrm{vac}} / U_{\mathrm{Coul}}^{\mathrm{dry}(4)}$ & 4.00 & 4.00 \\
$U_{\mathrm{Coul}}(3\;\mathrm{\AA}) / U_{\mathrm{Coul}}(6\;\mathrm{\AA})$ & 2.00 & 2.00 \\
Neutral Ar: $U_{\mathrm{vdW}}^{\mathrm{vac}} = U_{\mathrm{vdW}}^{\mathrm{sol}}$ & exact & exact \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Phase~14 status: PASS.}  24 checks, 0 failures.

%% ====================================================================

\section{Cumulative Audit Summary}

\begin{center}
\begin{tabular}{lrl}
\toprule
Phase & Checks & Status \\
\midrule
1 (kernel)     & 11 & PASS \\
2 (structural) & 15 & PASS \\
3 (relaxation) & 17 & PASS \\
4 (presets)    & 52 & PASS \\
5 (crystals)   & 13 & PASS \\
6/7 (sessions) & 43 & PASS \\
8--10 (env)    & 28 & PASS \\
11--13 (comp)  & 26 & PASS \\
14 (milestones)& 24 & PASS \\
\midrule
\textbf{Total} & \textbf{229} & \textbf{ALL PASS} \\
\bottomrule
\end{tabular}
\end{center}

% ===========================================================================
\newpage
\section{Verification Milestone A --- Deterministic Baseline Kernel Validation}
% ===========================================================================

\subsection{Status}

\begin{center}
\large\bfseries 256 / 256 \quad PASS \quad 0 FAIL \quad 0 SKIP
\end{center}

\noindent\textbf{Executable:} \texttt{apps/deep\_verification.cpp}\\
\textbf{Platform:} GNU/Linux x86\_64 (WSL2 6.6 / Debian), g++ 14.2, CMake 3.31\\
\textbf{Branch:} \texttt{2.5.2.X}

\subsection{Milestone Statement}

The deep verification framework now demonstrates full pass status for the
currently implemented deterministic baseline across Lennard-Jones interactions,
Coulomb electrostatics, dielectric screening, small-molecule bond detection,
reference crystal geometry, composite force-energy consistency, noble-gas
cluster relaxation descent, randomised supercell replication, and NVE energy
conservation. Randomised seeded runs further confirm reproducibility and local
consistency of force-energy behaviour. External empirical-reference suites pass
when downloaded reference data are provisioned, and otherwise skip cleanly
without indicating model failure.

\medskip\noindent\textit{This result marks completion of baseline deterministic
kernel revalidation for the present physics scope, shifting the next development
stage from foundational correctness recovery toward convergence tightening,
expanded physical coverage, and formation-oriented workflow integration.}

\subsection{Suite Summary}

\begin{center}
\begin{tabular}{clrr}
\toprule
Suite & Description & Checks & Result \\
\midrule
A & LJ parameter audit (42 UFF elements, Rapp\'{e} 1992) & 42 & PASS \\
B & Homonuclear dimer sweeps & 20 & PASS \\
C & Ion-pair Coulomb + force consistency & 15 & PASS \\
D & Crystal geometry vs Wyckoff (30 structures) & 90 & PASS \\
E & Molecule bond detection (9 molecules) & 9 & PASS \\
F & Composite force-energy (H\textsubscript{2}O, NH\textsubscript{3}, CH\textsubscript{4}, Ar\textsubscript{2}) & 4 & PASS \\
G & Noble-gas cluster FIRE, two-layer (Ar\textsubscript{2}--Ar\textsubscript{7}) & 30 & PASS \\
H & Dielectric sweep (19 CRC 104 solvents) & 19 & PASS \\
N & Restoring-force scan: He\textsubscript{2}--Xe\textsubscript{2} & 10 & PASS \\
Q & Crystal builder fidelity, 4 presets $2\times2\times2$ & 8 & PASS \\
I & Randomised LJ force-energy ($N=500$, seeded) & 1 & PASS \\
J & Randomised Coulomb $1/r$ ($N=500$, seeded) & 2 & PASS \\
K & Randomised dielectric ratio ($N=500$, seeded) & 1 & PASS \\
L & Randomised supercell atom count & 6 & PASS \\
M & NVE energy conservation (Ar\textsubscript{3}--Ar\textsubscript{5}) & 6 & PASS \\
P & Downloaded empirical refs (PubChem + NIST) & 53 & PASS \\
\midrule
\textbf{Total} & & \textbf{256} & \textbf{ALL PASS} \\
\bottomrule
\end{tabular}
\end{center}

\noindent Canonical seed: \texttt{1772952709}\quad
Reproduce: \texttt{./deep\_verification --seed 1772952709 --rand-iters 500}

\subsection{Selected Metrics}

\subsubsection*{Suite G --- FIRE Convergence}

\begin{center}
\begin{tabular}{lrrrr}
\toprule
Cluster & $U_0$ (kcal/mol) & $U_f$ & $F_\mathrm{rms}$ (kcal/mol/\AA) & Steps \\
\midrule
Ar\textsubscript{2} & $-0.026$ & $-0.238$ & $9.98\times10^{-6}$ & 603 \\
Ar\textsubscript{3} & $-0.696$ & $-0.714$ & $5.35\times10^{-6}$ & 308 \\
Ar\textsubscript{4} & $-1.219$ & $-1.428$ & $4.78\times10^{-6}$ & 919 \\
Ar\textsubscript{5} & $-1.675$ & $-2.167$ & $8.72\times10^{-6}$ & 1269 \\
Ar\textsubscript{6} & $-2.077$ & $-3.025$ & $4.97\times10^{-6}$ & 2075 \\
Ar\textsubscript{7} & $-2.462$ & $-3.928$ & $8.55\times10^{-6}$ & 2491 \\
\bottomrule
\end{tabular}
\end{center}

\subsubsection*{Suite M --- NVE Energy Drift}

\begin{center}
\begin{tabular}{lrr}
\toprule
System & $E_0$ (kcal/mol) & Max drift \\
\midrule
Ar\textsubscript{3} & 1.167 & $2.91\times10^{-12}$ \\
Ar\textsubscript{4} & 1.817 & $5.27\times10^{-12}$ \\
Ar\textsubscript{5} & 0.314 & $2.26\times10^{-11}$ \\
\bottomrule
\end{tabular}
\end{center}

\subsection{FIRE Implementation Correction}

A deadlock in the Euler-FIRE integrator was identified and corrected.
After a $P < 0$ velocity reset in the pure-Euler formulation
($\mathbf{X} \leftarrow \mathbf{X} + \mathbf{V}\,\Delta t$, no acceleration
term), setting $\mathbf{V} = \mathbf{0}$ caused $\Delta U = 0$ on the next
evaluation, falsely triggering the $\varepsilon_U$ stop criterion.
The correction: after $P < 0$ reset, velocities are kicked along the
normalised force direction with magnitude $\Delta t$, and the $\varepsilon_U$
check is conditioned on $\varepsilon_U > 0$.

\subsection{Known Limitations}

\begin{itemize}
\item UFF LJ $r_\mathrm{min}$ exceeds actual bond lengths for Cu, Fe, Si;
  FIRE relaxation of metallic/covalent crystals is not applicable.
  Suite Q tests crystal builder geometry only.
\item Suite G tests descent from a single Fibonacci-sphere start; global-minimum
  search for LJ clusters is a future planned suite.
\item Bonded harmonic terms (stretch, angle) present in the codebase are not
  yet covered by \texttt{deep\_verification}.
\item PBC is not exercised in Suites G, N, or Q; PBC stability is a next target.
\end{itemize}

\end{document}
